---
title: "Bike Rides Analysis"
author: "Pooja Rathee"
date: '2022-04-08'
output:
  pdf_document: default
  html_document: default
---

# Load Libraries.

```{r echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

Read all data files separately.

```{r}
dt1 <- read_csv("data/202104-divvy-tripdata.csv")
dt2 <- read_csv("data/202105-divvy-tripdata.csv")
dt3 <- read_csv("data/202106-divvy-tripdata.csv")
dt4 <- read_csv("data/202107-divvy-tripdata.csv")
dt5 <- read_csv("data/202108-divvy-tripdata.csv")
dt6 <- read_csv("data/202109-divvy-tripdata.csv")
dt7 <- read_csv("data/202110-divvy-tripdata.csv")
dt8 <- read_csv("data/202111-divvy-tripdata.csv")
dt9 <- read_csv("data/202112-divvy-tripdata.csv")
dt10 <- read_csv("data/202201-divvy-tripdata.csv")
dt11 <- read_csv("data/202202-divvy-tripdata.csv")
dt12 <- read_csv("data/202203-divvy-tripdata.csv")

```

# Combine All files into `all_trips`

```{r}
all_trips <- rbind(dt1, dt2, dt3, dt4, dt5, dt6, dt7, dt8, dt9, dt10, dt11, dt12)

# Convert ride_id and rideable_type to character so that they can stack correctly
all_trips <-  mutate(all_trips, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
colnames(all_trips)
```
# Inspect the new table that has been created
```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

# Examine the table
```{r}
table(all_trips$member_casual)
```

# Formatting the data
```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```                     
                           

# Add a "ride_length" calculation to all_trips (in seconds)
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

# Inspect the structure of the columns
str(all_trips)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

# Reassign the data
```{r}
all_trips_v2 <- all_trips[!( all_trips$ride_length<0),]
```

## Find the mean, median, max and min
```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

## summarize the data
```{r}
summary(all_trips_v2$ride_length)
```

# Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

# See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

# Notice that the days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
# Now, let's run the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
# analyze ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by member_casual and weekday
  summarise(number_of_rides = n()						          	#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 	         	# calculates the average duration
  arrange(member_casual, weekday)							        	# sorts
```

# Let's create a visualization for average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
labs(title = "Relationship between average_duration and weekday")


```
# Average duration of rides generally increases on Sunday and Saturday as compare to weekdays.
